Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-04 appendix.tex
answering EMs

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.

@@

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
Similarly we enumerate ancestral splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

We first fix notation.
\begin{definition}
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$.
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \alignmentColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\alignmentColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet)
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \ancestralStateColumn  & \mbox{if } \alignmentColumn\in\mathcal{P}(\siteSplitSet) \\
        \overline{\ancestralStateColumn}  & \mbox{if } \alignmentColumn\notin\mathcal{P}(\siteSplitSet).
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
that takes corresponding realizations $\siteSplit_j$ for some $j$, and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
\]
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the complement of a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if it is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if a set of tip labels $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\ancestralStateColumn$; otherwise, if $\alignmentColumn$ is not in $\siteSplitSet$, then the complement of $\alignmentColumn$ necessarily is in $\siteSplitSet$, and $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the complement of $\ancestralStateColumn$.

For the $i$th factor of \eqref{eq:full_likelihood},
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it maps all possible values from $\alphabet^\nAncestralStateRows$ to an element in $\mathcal{P}(\ancestralSplitSet)$, and from this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ depending on $i$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
\]

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralStateColumn=\emptyset\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralStateColumn=\{1\}\mid \alignmentColumn=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
For the binary symmetric model $\alphabet=\{0,1\}$.
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
We define an ``ancestral state split'' in a similar way, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$
In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
%das: done

@@

We enumerate the site splits $\siteSplit_j$ of which there are $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ in total where $\mathcal{P}$ denotes the power set.
%EM I think that we should either use "ancestral splits" everywhere, or "ancestral state splits."
%das: yah, agreed
Similarly we enumerate ancestral state splits $\ancestralSplit_k$ of which there are $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$ in total.

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nSiteRows\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
\]
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\siteSplitSet$.

For the $i$th factor of \eqref{eq:full_likelihood},
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,i'}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,i'}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ depending on $i$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral state splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral state splits for $\siteSplit_1=\emptyset$ are
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
\]
\[
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
\]
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
\[
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
\]
For $\siteSplit_5=\{1,2\}$ we have
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
\]
\[
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
\]
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
\[
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
\]

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
Using branch fidelities, the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the InvFels tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} (for the InvFels tree) and~\ref{tab:fels_likelihoods} (for the Felsenstein tree).
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like00}
\caption[short]{}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_like10}
\caption[short]{}
\end{subfigure}
\caption{
    Example likelihood computations on the InvFels tree $\tau_1$ for fidelities $x$, $y$, and $w$.
    Edges labeled by the probability of substitution along that edge.
    In (a), we compute the product to obtain $\Pr(\ancestralSplitRV=\emptyset\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1+w)/32$.
    In (b), the same process yields $\Pr(\ancestralSplitRV=\{1\}\mid \siteSplitRV=\{2,3\},\tau_1,t) = (1+x)(1-x)(1+y)(1-y)(1-w)/32$.
}
\label{fig:example_likelihoods}
\end{figure}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}
